### 2022-10-26 [2020-08]_ACMTOMS_(MQSI)/paper/TeX/2-monotone-quintic.tex
Moved all paper files into a TeX directory. Creating LaTeX version next.

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large local $L^2$ norm
(approximations with large {\it wiggle}). In an attempt to capture the
local shape of the data while minimizing wiggle, this package uses a
facet model from image processing [Haralick and Watson 1981] to
estimate first and second derivatives at breakpoints. Rather than
picking a local linear or quadratic fit with minimal residual, this
work \red{instead} % uses a quadratic facet model that
selects the local quadratic
interpolant with minimum magnitude second derivative \red{(hereafter
referred to as a {\it quadratic facet model}). The term ``facet'' 
refers to one piece of an image surface modelled by a two-dimensional
piecewise polynomial.}.

@@

Given \red{that} the \red{initial derivative estimates locally minimize 
the magnitude of the second derivative}, it is desirable to make
the smallest necessary changes to the initial interpolating spline $Q$
while enforcing monotonicity. In practice a binary search for the
boundary of monotonicity is used in place of solely shrinking $DQ$ and
$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
over which $Q$ is nonmonotone at least once during the search. The
binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
1$, that is true if the order six polynomial piece of $Q(x)$ on
$[X_i, X_{i+1}]$ matching derivatives
$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false
otherwise.  The binary search is only applied at those breakpoints
adjoining intervals $[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and
hence $B_i(1)$ is false.  It is further assumed that there exists
$0 \le s^* \le 1$ such that $B_i(s)$ is true for $0 \le s \le s^*$ and
false for some $1 > s > s^*$. Since the derivative conditions at
interior breakpoints are shared by intervals left and right of the
breakpoint, the binary search is performed at all breakpoints
simultaneously.  Specifically, the monotonicity of $Q$ is checked on
all active intervals in each step of the binary search to determine
the next derivative modification at each breakpoint. The goal of this
search is to converge on the boundary of the monotone region in the
$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
Watson [1994]) for all intervals. This multiple-interval binary search
allows the value zero to be obtained for all (first and second)
derivative values in a fixed maximum number of computations, hence has
no effect on computational complexity order. This binary search
algorithm is outlined below.

>>>>

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations whose
second derivative functions often have large local $L^2$ norm
(approximations with large {\it wiggle}). In an attempt to capture the
local shape of the data while minimizing wiggle, this package uses a
facet model from image processing [Haralick and Watson 1981] to
estimate first and second derivatives at breakpoints. Rather than
picking a local linear or quadratic fit with minimal residual, this
work instead selects the local quadratic interpolant with minimum
magnitude second derivative (hereafter referred to as a {\it quadratic
  facet model}). The term ``facet'' refers to one piece of an image
surface modelled by a two-dimensional piecewise polynomial.

@@

Given that the initial derivative estimates locally minimize 
the magnitude of the second derivative, it is desirable to make
the smallest necessary changes to the initial interpolating spline $Q$
while enforcing monotonicity. In practice a binary search for the
boundary of monotonicity is used in place of solely shrinking $DQ$ and
$D^2Q$ at breakpoints adjoining {\it active\/} intervals: intervals
over which $Q$ is nonmonotone at least once during the search. The
binary search considers a Boolean function $B_i(s)$, for $0 \le s \le
1$, that is true if the order six polynomial piece of $Q(x)$ on
$[X_i, X_{i+1}]$ matching derivatives
$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and 
derivatives $Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, 
$D^2Q(X_{i+1})=s \,v_{i+1}$ at $X_{i+1}$ is monotone, and false
otherwise.  The binary search is only applied at those breakpoints
adjoining intervals $[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and
hence $B_i(1)$ is false.  It is further assumed that there exists
$0 \le s^* \le 1$ such that $B_i(s)$ is true for $0 \le s \le s^*$ and
false for some $1 > s > s^*$. Since the derivative conditions at
interior breakpoints are shared by intervals left and right of the
breakpoint, the binary search is performed at all breakpoints
simultaneously.  Specifically, the monotonicity of $Q$ is checked on
all active intervals in each step of the binary search to determine
the next derivative modification at each breakpoint. The goal of this
search is to converge on the boundary of the monotone region in the
$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
Watson [1994]) for all intervals. This multiple-interval binary search
allows the value zero to be obtained for all (first and second)
derivative values in a fixed maximum number of computations, hence has
no effect on computational complexity order. This binary search
algorithm is outlined below.
